Statistical Model Selection for Markov Models of Biomolecular Dynamics

### 2014-04-15 paper/manuscript/manuscript.tex
added manuscript dir and also fixed formatting things

\title{Statistical Model Selection for Markov Models of Biomolecular Dynamics}
\author{Christian R. Schwantes}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Robert T. McGibbon}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Vijay S. Pande}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Biophysics Program, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Computer Science, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Structural Biology, Stanford University, Stanford CA 94305, USA}
\email{pande@stanford.edu}
\keywords{MSM, molecular dynamics, AIC, BIC, Bayes factor}

@@

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly. Application of techniques that consider both systematic bias and statistical error on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those timescale-based techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markov state model construction.
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs_final/mull_ww.png}
\caption{Systems studied in this work. (A) Brownian dynamics on the two-dimensional M\"{u}ller potential \cite{Muller1979Location}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\label{fig:pics}
\end{figure}

Two systems were investigated using this likelihood scheme. The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential. The dynamics are governed by
$$
\frac{d\mathbf{x}}{dt} = - \nabla V(\mathbf{x})\zeta + \sqrt{2kT\zeta} R(t),
$$ where $\zeta = 10^{-3}$, $kT = 15$, $R(t)$ is a delta-correlated Gaussian process with zero mean, and $V(\mathbf{x})$ was defined as
$$
V(\mathbf{x}) =  \sum_{j=1}^4 A_j \cdot \exp\Big(a_j (x_1-X_j)^2+b_j(x_1-X_j)(x_2-Y_j)+c_j(x_2-Y_j)^2\Big),
$$ where $a = (-1, -1, -6.5, 0.7)$; $b = (0, 0, 11, 0.6)$; $c = (-10, -10, -6.5, 0.7)$; $A = (-200, -100, -170, 15)$; $X = (1, 0, -0.5, -1)$; $Y = (0, 0.5, 1.5, 1)$ as suggested by \citet{Muller1979Location}. Using the Euler-Maruyama method and a time step of $0.1$, we produced two trajectories of length $10^6$ time steps. The initial positions were sampled via a uniform distribution over the box: $[-1.5, 2.0] \times [-0.2, 2.0]$.

@@

How well can likelihood-based methods select parameters for an MSM in the data-rich regime? To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see \emph{Methods}). We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence. The analysis \cref{fig:mullerlike} indicates that the log-likelihood on the training data set continually increases with the addition of further states. The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small. For this data set, the log-likelihoods computed on a separate test set exhibit a notably similar trend. Above $k=600$, the test log-likelihood stops increasing. These two trends are to be expected: the additional parameters in the model at higher $k$ never decreases the model's ability to fit the training data. On the other hand, this can lead to the model fitting the statistical noise in the training data set, rather than the system's true dynamics. As result, these "over fit" models at high $k$ are less predictive, as measured by the log-likelihood they assign to new test data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=6in]{figs/mull/mull.pdf}
\caption{Four model selection criteria were used (left panel) to select the state discretization for MSMs built on simulations of the M\"{u}ller potential. The log-likelihood was evaluated on the training set (black) and a separate test set (blue). These values are plotted as $-2\cdot \log L$ for ease of comparison with the BIC (red) and AIC (purple). Lower scores indicate better models. Dashed lines are used to indicate criterion-minimizing models. These criteria are consistent with the convergence of the implied timescales (right panel).}
\label{fig:mullerlike}
\end{figure}

@@

\begin{figure}
\centering
\label{fig:ww-tica}
\includegraphics[width=6in]{figs/ww_tica/ww_tica.pdf}

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_pca/ww_pca.pdf}
\caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA. The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states. The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.\label{fig:ww-pca}} 
\end{figure}

>>>>

\setkeys{acs}{usetitle=true}

\title{Statistical Model Selection for Markov Models of Biomolecular Dynamics}
\author{Robert T. McGibbon}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Christian R. Schwantes}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\altaffiliation{Contributed equally to this work}
\author{Vijay S. Pande}
\affiliation{Department of Chemistry, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Biophysics Program, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Computer Science, Stanford University, Stanford CA 94305, USA}
\alsoaffiliation{Department of Structural Biology, Stanford University, Stanford CA 94305, USA}
\email{pande@stanford.edu}

@@

\begin{document}
\maketitle
\begin{abstract}
Markov state models provide a powerful framework for the analysis of biomolecular conformation dynamics in terms of their metastable states and transition rates. These models provide both a quantitative and comprehensible description of the long-timescale dynamics of large molecular dynamics with a Master equation, and have been successfully used to study protein folding, protein conformational change, and protein-ligand binding. However, to achieve satisfactory performance, existing methodologies often require expert intervention when defining the model's discrete state space. While standard model selection methodologies focus on the minimization of systematic bias and disregard statistical error, we show that by consideration of the states' conditional distribution over conformations, both sources of error can be balanced evenhandedly. Application of techniques that consider both systematic bias and statistical error on two 100$\mu s$ molecular dynamics trajectories of the Fip35 WW domain shows agreement with existing techniques based on self-consistency of the model's relaxation timescales, with more suitable results in regimes in which those timescale-based techniques encourage over-fitting. By removing the need for expert tuning, these methods should reduce modeling bias and lower the barriers to entry in Markov state model construction.

\vspace{1em}
\noindent KEYWORDS: MSM, molecular dynamics, AIC, BIC, Bayes factor, model selection
\end{abstract}

@@

\begin{figure}
\centering
\includegraphics[width=3in]{figs/mull_ww.png}
\caption{Systems studied in this work. (A) Brownian dynamics on the two-dimensional M\"{u}ller potential \cite{Muller1979Location}. (b) 200 $\mu s$ of dynamics of the Fip35 WW domain\cite{Liu2008Experimental}, courtesy of D.E. Shaw research \cite{Shaw2010Atomic}.}
\label{fig:pics}
\end{figure}

Two systems were investigated using this likelihood scheme (\cref{fig:pics}). The first was a simple two dimensional surface with three energy minima, called the M\"{u}ller potential. The dynamics are governed by
$$
\frac{d\mathbf{x}}{dt} = - \nabla V(\mathbf{x})\zeta + \sqrt{2kT\zeta} R(t),
$$ where $\zeta = 10^{-3}$, $kT = 15$, $R(t)$ is a delta-correlated Gaussian process with zero mean, and $V(\mathbf{x})$ was defined as
$$
V(\mathbf{x}) =  \sum_{j=1}^4 A_j \cdot \exp\Big(a_j (x_1-X_j)^2+b_j(x_1-X_j)(x_2-Y_j)+c_j(x_2-Y_j)^2\Big),
$$ where $a = (-1, -1, -6.5, 0.7)$; $b = (0, 0, 11, 0.6)$; $c = (-10, -10, -6.5, 0.7)$; $A = (-200, -100, -170, 15)$; $X = (1, 0, -0.5, -1)$; $Y = (0, 0.5, 1.5, 1)$ as suggested by \citet{Muller1979Location}. Using the Euler-Maruyama method and a time step of $0.1$, we produced two trajectories of length $10^6$ time steps. The initial positions were sampled via a uniform distribution over the box: $[-1.5, 2.0] \times [-0.2, 2.0]$.

@@

How well can likelihood-based methods select parameters for an MSM in the data-rich regime? To answer this question, we simulated a simple two-dimensional potential and built a series of MSMs (see \emph{Methods}). We compared four different model selection criteria: (1) test log-likelihood (2) BIC (3) AIC and (4) implied timescales convergence. The analysis (\cref{fig:mullerlike}) indicates that the log-likelihood on the training data set continually increases with the addition of further states. The increase is most dramatic at low $k$, and levels off at high $k$, indicating that once $k$ is large enough, the marginal increase in the likelihood with respect to further states is small. For this data set, the log-likelihoods computed on a separate test set exhibit a notably similar trend. Above $k=600$, the test log-likelihood stops increasing. These two trends are to be expected: the additional parameters in the model at higher $k$ never decreases the model's ability to fit the training data. On the other hand, this can lead to the model fitting the statistical noise in the training data set, rather than the system's true dynamics. As result, these "over fit" models at high $k$ are less predictive, as measured by the log-likelihood they assign to new test data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=6in]{figs/mull.pdf}
\caption{Four model selection criteria were used (left panel) to select the state discretization for MSMs built on simulations of the M\"{u}ller potential. The log-likelihood was evaluated on the training set (black) and a separate test set (blue). These values are plotted as $-2\cdot \log L$ for ease of comparison with the BIC (red) and AIC (purple). Lower scores indicate better models. Dashed lines are used to indicate criterion-minimizing models. These criteria are consistent with the convergence of the implied timescales (right panel).}
\label{fig:mullerlike}
\end{figure}

@@

\begin{figure}
\centering
\label{fig:ww-tica}
\includegraphics[width=6in]{figs/ww_tica.pdf}

@@

\begin{figure}
\centering
\includegraphics[width=6in]{figs/ww_pca.pdf}
\caption{The four selection criteria were used to compare MSMs for the Fip35WW domain built with PCA. The longest relaxation timescale (right panel) exhibits a linear increase up to 25,000 states. The AIC (blue), BIC (red) and test set log-likelihood each recommend a model with 200 states.\label{fig:ww-pca}} 
\end{figure}
